Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-11 paper/main.tex
discussion draft

@@

Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manual and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are ``whitelists'' or 
``blacklists'' of edges 
\citep{scutari2010}, or specification
of a ``tiered'' (e.g., temporal) structure between the variables \citep{Bang2023}.
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty 
of our method lies in the interactive
back-and-forth between domain knowledge and data, 
which provably leads to the correct
result in a polynomial amount of steps in the oracle settings we considered.

>>>>

It is possible to integrate domain knowledge into score-based methods
in a similar way as we do here for constraint-based structure learning.
\citet{Kitson_2025} recently extended the Tabu search algorithm by a procedure
that asks a domain expert for advice before making changes that are classified as 
``uncertain'' such as in cases of tied score improvement or low-confidence modifications. 
In contrast, our approach places greater emphasis on expert input as no edges are oriented
without consulting the expert. Further, the approach by \citet{Kitson_2025} 
requires experts that are able to distinguish between direct and indirect effects.

@@

Researchers often prefer to construct DAGs manually rather than using automated
algorithms, which could be partly due to practical challenges in applying automated methods.
To assist this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manual and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are ``whitelists'' or 
``blacklists'' of edges 
\citep{scutari2010}, or specification of a ``tiered'' (e.g., temporal) 
structure between the variables \citep{Bang2023}.
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty 
of our method lies in the interactive
back-and-forth between domain knowledge and data, 
which provably leads to the correct
result in a polynomial amount of steps in the oracle settings we considered.

Nevertheless, the algorithm presented in this paper has 
significant limitations. First of all, the causal Markov and causal sufficiency assumptions are
widely seen as too restrictive, since they do not allow for latent confounders.
It would therefore be interesting to extend our approach to a more general 
algorithm such as FCI \cite{Spirtes1995}. Further, while it is theoretically possible
to recover from some expert mistakes that introduce cycles, this requires searching for a 
separating set for each edge on each cycle, which in the worst case can take exponential time 
and would then essentially amount to running significant parts of the PC algorithm. 
In practice, we may be better off breaking cycles by other means -- such as the heuristics
implemented in our experiments -- even if the theoretical correctness guarantee is lost. 
Lastly, any iterative model improvement procedure runs the risk of overfitting to the 
given dataset. While this is also the case for purely data-driven structure learning
(which may even require a vastly larger amount of model improvement steps), the 
inclusion of human judgement in this process means that it will be seen as less reliable.
Models generated by this approach should certainly be seen as preliminary and in need
of further validation using independent data, and ideally a full log of all human decisions 
made during the process should be kept. In this respect, it will be interesting to see how
the research community will feel about the use of LLMs instead of humans for model refinement.
